Lessons learned and paths forward for rabies dog vaccination in Madagascar: a case study of pilot vaccination campaigns in Moramanga District

### 2021-02-20 paper/mdpi_paper.tex
adding subfiles for trop med

%=================================================================
\documentclass[water,article,submit,moreauthors,pdftex]{mdpi}

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{Full title of the paper (Capitalized)}

% Authors, for the paper (add full first names)
\Author{Dominik
Leutnant$^{1,2,\ddagger,*}$\href{https://orcid.org/0000-0003-3293-2315}{\orcidicon}, John
Doe$^{2, \dagger, \ddagger}$}

% Authors, for metadata in PDF
\AuthorNames{Dominik Leutnant, John Doe}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Muenster University of Applied Sciences - Institute for
Infrastructure, Water, Resources, Environment Correnstr. 25, 48149
Muenster,
Germany; \href{mailto:leutnant@fh-muenster.de}{\nolinkurl{leutnant@fh-muenster.de}}\\
$^{2}$ \quad Your department Street, City,
Country; \href{mailto:mail@mail.com}{\nolinkurl{mail@mail.com}}\\
}
% Contact information of the corresponding author
\corres{Correspondence: \href{mailto:leutnant@fh-muenster.de}{\nolinkurl{leutnant@fh-muenster.de}};
Tel.: +XX-000-00-0000.}

% Current address and/or shared authorship
\firstnote{Current address: Updated affiliation}
\secondnote{These authors contributed equally to this work.}

@@

% Simple summary
\simplesumm{A Simple summary goes here.}

@@

% Keywords
\keyword{keyword 1; keyword 2; keyword 3 (list three to ten pertinent
keywords specific to the article, yet reasonably common within the
subject discipline.).}

@@

@@

\hypertarget{version}{%
\section{Version}\label{version}}

This Rmd-skeleton uses the mdpi Latex template published 2019/02.
However, the official template gets more frequently updated than the
`rticles' package. Therefore, please make sure prior to paper
submission, that you're using the most recent .cls, .tex and .bst files
(available \href{http://www.mdpi.com/authors/latex}{here}).

@@

The introduction should briefly place the study in a broad context and
highlight why it is important. It should define the purpose of the work
and its significance. The current state of the research field should be
reviewed carefully and key publications cited. Please highlight
controversial and diverging hypotheses when necessary. Finally, briefly
mention the main aim of the work and highlight the principal
conclusions. As far as possible, please keep the introduction
comprehensible to scientists outside your particular field of research.
Citing a journal paper
\citep{bertrand-krajewski_distribution_1998, leutnant_stormwater_2016}.
And now citing a book reference \citet{gujer_systems_2008}. Please use
the command \citep{leutnant_stormwater_2016} for the following MDPI
journals, which use author-date citation: Administrative Sciences, Arts,
Econometrics, Economies, Genealogy, Humanities, IJFS, JRFM, Laws,
Religions, Risks, Social Sciences.

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

Materials and Methods should be described with sufficient details to
allow others to replicate and build on published results. Please note
that publication of your manuscript implicates that you must make all
materials, data, computer code, and protocols associated with the
publication available to readers. Please disclose at the submission
stage any restrictions on the availability of materials or information.
New methods and protocols should be described in detail while
well-established methods can be briefly described and appropriately
cited.

Research manuscripts reporting large datasets that are deposited in a
publicly available database should specify where the data have been
deposited and provide the relevant accession numbers. If the accession
numbers have not yet been obtained at the time of submission, please
state that they will be provided during review. They must be provided
prior to publication.

Interventionary studies involving animals or humans, and other studies
require ethical approval must list the authority that provided approval
and the corresponding ethical approval code.

@@

This section may be divided by subheadings. It should provide a concise
and precise description of the experimental results, their
interpretation as well as the experimental conclusions that can be
drawn.

\hypertarget{subsection-heading-here}{%
\subsection{Subsection Heading Here}\label{subsection-heading-here}}

Subsection text here.

\hypertarget{subsubsection-heading-here}{%
\subsubsection{Subsubsection Heading
Here}\label{subsubsection-heading-here}}

Bulleted lists look like this:

\begin{itemize}
\tightlist
\item
  First bullet
\item
  Second bullet
\item
  Third bullet
\end{itemize}

Numbered lists can be added as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  First item
\item
  Second item
\item
  Third item
\end{enumerate}

The text continues here.

All figures and tables should be cited in the main text as Figure 1,
Table 1, etc.

\begin{figure}[H]
\centering
\includegraphics[width=3 cm]{logo-mdpi}
\caption{This is a figure, Schemes follow the same formatting. If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel. (\textbf{b}) Description of what is contained in the second panel. Figures should be placed in the main text near to the first time they are cited. A caption on a single line should be centered.}
\end{figure}

\begin{table}[H]
\caption{This is a table caption. Tables should be placed in the main text near to the first time they are cited.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Title 1}    & \textbf{Title 2}  & \textbf{Title 3}\\
\midrule
entry 1     & data          & data\\
entry 2     & data          & data\\
\bottomrule
\end{tabular}
\end{table}

This is an example of an equation:

\begin{equation}
\mathbb{S}
\end{equation}

Example of a theorem:

\begin{Theorem}
Example text of a theorem.
\end{Theorem}

The text continues here. Proofs must be formatted as follows:

Example of a proof:

\begin{proof}[Proof of Theorem 1]
Text of the proof. Note that the phrase `of Theorem 1' is optional if it is clear which theorem is being referred to.
\end{proof}

The text continues here.

>>>>

%=================================================================
\documentclass[tropicalmed,article,submit,moreauthors,pdftex]{mdpi}

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{Lessons learned and paths forward for rabies dog vaccination in
Madagascar: a case study of pilot vaccination campaigns in Moramanga
District}

% Authors, for the paper (add full first names)
\Author{Caitlynn Filla$^{1, \dagger}$, Malavika Rajeev$^{2,
\dagger}$\href{https://orcid.org/0000-0001-6030-4995}{\orcidicon}, Zoavina
Randriana$^{3}$, Chantal Hanitriniana$^{4}$, Radoniaina R.
Rafaliarison$^{3}$, Glenn Edosoa$^{5}$, Mamitiana
Andriamananjara$^{6}$, Nivohanitra P. Razafindraibe$^{6}$, José
Nely$^{7}$, Angelique Ferreira$^{3, 8}$, Annie Yang$^{2,
\ddagger}$, Fenomanana Daniel$^{3}$, Tara A. Clarke$^{3, 9}$, Zachary
Farris$^{3, 10}$, Terry Stone$^{8}$, Jochem Lastdrager$^{8}$, Tsiky
Rajaonarivelo$^{3}$, Katie Hampson$^{11}$, C. Jessica E.
Metcalf$^{2}$, Kim Valenta$^{1, 3}$}

% Authors, for metadata in PDF
\AuthorNames{Caitlynn Filla, Malavika Rajeev, Zoavina Randriana, Chantal
Hanitriniana, Radoniaina R. Rafaliarison, Glenn Edosoa, Mamitiana
Andriamananjara, Nivohanitra P. Razafindraibe, José Nely, Angelique
Ferreira, Annie Yang, Fenomanana Daniel, Tara A. Clarke, Zachary
Farris, Terry Stone, Jochem Lastdrager, Tsiky Rajaonarivelo, Katie
Hampson, C. Jessica E. Metcalf, Kim Valenta}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad University of Florida, Department of Anthropology
Gainesville, Florida, USA; \\
$^{2}$ \quad Department of Ecology and Evolutionary Biology, Princeton
University Princeton, NJ, United States; \\
$^{3}$ \quad The Mad Dog Initiative Akanin'ny Veterinera, Akaikiniarivo,
Ambatobe, Antananarivo 101, Madagascar; \\
$^{4}$ \quad Mention Zoologie et Biodiversité Animale, Faculté des
Sciences Université d'Antananarivo, Antananarivo, Madagascar; \\
$^{5}$ \quad Chargé des Maladies Tropicales Négligées Organisation
mondiale de la Santé Madagascar, Antananarivo, Madagascar; \\
$^{6}$ \quad Direction des Services Vétérinaires Ministère chargé de
l'Agriculture et de l'Élevage, Antananarivo, Madagascar; \\
$^{7}$ \quad Service contre les Maladies Endémo-épidémiques et
Tropicales Négligées Ministère de la Santé Publique, Antananarivo,
Madagascar; \\
$^{8}$ \quad Travelling Animal Doctors 919 Crossan Rd, Newark, Delaware,
USA; \\
$^{9}$ \quad North Carolina State University, Department of Sociology
and Anthropology Raleigh, North Carolina, USA; \\
$^{10}$ \quad Appalachian State University, Department of Health and
Exercise Science Boone, North Carolina, USA; \\
$^{11}$ \quad Boyd Orr Centre for Population and Ecosystem Health
Institute of Biodiversity, Animal Health and Comparative Medicine
University of Glasgow, Glasgow, UK; \\
}
% Contact information of the corresponding author
\corres{Correspondence: \href{mailto:mrajeev@princeton.edu}{\nolinkurl{mrajeev@princeton.edu}}}

% Current address and/or shared authorship
\firstnote{These authors contributed equally to this work.}
\secondnote{Deceased.}

@@

% Simple summary

@@

% Keywords
\keyword{canine rabies; mass dog vaccination; central point vaccination;
puppy vaccination; Zeroby30;}

@@

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}

@@

@@

Canine rabies results in an estimated 60,000 human deaths per year
globally \citep{hampson2015}. These deaths are entirely preventable:
prompt post-exposure prophylaxis of humans exposed to rabies is highly
effective at preventing death and mass dog vaccination can interrupt
transmission in domestic dogs and eventually lead to disease elimination
\citep{fooks2014}. The World Health Organization (WHO) and its partners
have set a goal to eliminate human deaths due to canine rabies by the
year 2030 (`ZeroBy30')\citep{abela-ridder2016}. Annual dog vaccination
campaigns achieving at least 70\% coverage are the recommended target
for controlling rabies in domestic dog populations
\citep{worldhealthorganization2013}. However, achieving this coverage
target in low and middle income countries where the burden of human
rabies is highest can be challenging due to economic, ecological,
sociocultural, and political barriers \citep{fahrion2017}.

In sub-Saharan African countries, parenteral vaccinations implemented
through static point campaigns have been shown to be cost-effective and
feasible \citep{borse2018}. While most dogs in these settings are
considered free roaming, they are mostly owned and are accessible for
vaccination through these campaigns \citep{morters2014, jibat2015}.
However, reaching high coverage and maintaining vaccination campaigns at
scale requires sustained investment and coordination, and the challenges
in implementation largely reflect financial and logistical constraints
more than the feasibility of vaccination itself \citep{fahrion2017}.
Developing clear and context-specific strategies and lowering costs and
resources needed could help spur the implementation and scaling up of
campaigns in these countries.

In Madagascar, canine rabies has been endemic for over a century, and
for most of that period, the Institut Pasteur de Madagascar has provided
post-exposure prophylaxis free-of-charge to bite patients in the country
\citep{reynes2011}. Currently, there are only 31 clinics where these
human vaccines are available, and there is minimal dog vaccination due
to high costs to owners and lack of vaccine availability
\citep{rajeev2018}. Recent studies have estimated a high burden of human
deaths, masked by weak surveillance across the country
\citep{rajeev2020, rajeev2018}.

The veterinary sector is largely private, with 204 veterinarians
employed in hybrid private/public employment as designated district
veterinarians. While dog vaccination is rare, livestock officers and
veterinarians work together to implement cattle vaccination campaigns
for anthrax on an annual basis as mandated by the government (owners are
charged a fee per animal for these vaccines which vary by location)
\citep{mondiale2017evaluation}. No routine mass dog vaccinations have
been conducted on the island, although a few pilot programs have begun
in recent years, largely implemented by NGO-government partnerships.

Here, we summarize lessons learned through the implementation of pilot
vaccination programs in the Moramanga District of Madagascar, where
previous work has shown high incidence of dog rabies cases and human
rabies exposures. In 2018 and 2019, we deployed two different
vaccination strategies. In 2018, we carried out a larger scale
volunteer-led pilot vaccination campaign in two communes (sub-district
level) using a static point strategy. In 2019/2020, we provided
vaccines, all necessary supplies for vaccine administration, and a
per-vaccine fee to the District Veterinarian to vaccinate animals as
part of routine services delivered continuously. We compare time, human
resources, costs, and coverage estimates between these campaigns, and
using a demographic and vaccination model we further explore different
vaccination strategies based on what we learned during implementation.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{study-area}{%
\subsection{Study Area}\label{study-area}}

The Moramanga District is located midway between the central highlands
and the east coast of Madagascar, at an average altitude of 936 m. It
comprises 21 communes, covering approximately 7150 km\textsuperscript{2}
with an approximate human population of 347,000\citep{instat2021}.
Previous work in the district has established a high burden of rabies
exposures and deaths despite the availability of post-exposure
prophylaxis at the district hospital \citep{rajeev2018}. While Moramanga
is relatively close to the capital city of Antananarivo
(\textasciitilde{} 3 hours by bus), within the district, travel times
between locations are highly variable, with much of the population
living in more rural areas with limited access to roads and
transportation \citep{rajeev2020}. Before 2018, there were limited
animal rabies vaccination services, with most animal vaccines available
in the urban commune of Moramanga Ville where owners were often charged
\textgreater{} 15,000 Ariary (\textasciitilde{} 4.28 USD) per vaccine
administered.

\hypertarget{campaign}{%
\subsection{2018 Campaign}\label{campaign}}

In 2018, two NGOs (the Mad Dog Initiative (MDI) and Traveling Animal
Doctors (TAD)) organized a pilot vaccination campaign in collaboration
with the Department of Veterinary Services, and the Ministry of Public
Health in the District of Moramanga. This campaign focused on two
communes in Moramanga, Moramanga Ville (the district center) and
Andasibe (a rural commune surrounding Andasibe National Park), where
previously high incidence of probable rabies exposures (Moramanga Ville)
and a high burden of deaths (Andasibe) had been recorded
\citep{rajeev2018}.

The campaign was planned as a series of static point vaccination
stations covering 1 - 3 fokontany (i.e., sub-communes) per day. A week
before the campaign dates, the vaccination team informed the chief of
the fokontany about the campaign and provided posters advertising the
date of the vaccine and that it would be available at no cost to owners.
During the campaign, we used Rabisin (10 mL vials with 1 mL per dose,
Boehringer Ingelheim) to vaccinate both dogs and cats presented that
were at least 1 month old. As part of the campaign, owners were surveyed
by vaccinators about how many dogs and cats they owned in total (split
by \textgreater{} 1 year vs.~\textless{} 1 year in order to avoid
language ambiguities that might result in excluding puppies and
kittens), as well as if their dogs were free roaming (no restrictions on
movement by the owner/`mirenyreny', tied/`mifatora', or
fenced/`mifefy'). Vaccinations were delivered at no cost to owners, but
as animal vaccination is generally thought of as a paid service in
Madagascar, owners were asked how much in Ariary they would be willing
to pay to have one animal vaccinated against rabies. For each animal
vaccinated, we recorded the species (cat or dog), sex, approximate age
in years, and whether the animal had been previously vaccinated.

To assess coverage, post-vaccination coverage surveys were conducted
according to previously established
methodology\citep{sambo2017, gibson2015}. All animals vaccinated were
concurrently marked with a colored, non-toxic, livestock crayon along
the top of the head or back. Between 4 - 6 PM when dogs are most active
\citep{sambo2017} and on the same day as the campaign, in each
vaccination location two transect surveys were done. Each team (two
people per team) walked a transect for 1 hour starting in opposite
directions and accompanied by a local guide to ensure that walking paths
did not overlap. Teams recorded any marked and unmarked dogs observed,
their roaming status (whether roaming, inside a fence, or tied), and
their approximate age (greater or less than 1 year of age).

\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig1}

\hypertarget{figure-1.-photos-from-the-2018-campaign.-topleft-advertisement-for-the-campaign-posted-on-the-door-of-the-fokontany-office-as-the-campaign-starts-topright-a-dog-post-vaccination-marked-with-a-crayon-bottomleft-a-basket-of-puppies-brought-for-vaccination-bottomright-a-line-of-owners-and-dogs-waiting-for-vaccination.}{%
\paragraph{\texorpdfstring{\textbf{Figure 1. Photos from the 2018
campaign}. \emph{Topleft: advertisement for the campaign posted on the
door of the fokontany office as the campaign starts; Topright: a dog
post-vaccination marked with a crayon; Bottomleft: a basket of puppies
brought for vaccination; Bottomright: a line of owners and dogs waiting
for
vaccination.}}{Figure 1. Photos from the 2018 campaign. Topleft: advertisement for the campaign posted on the door of the fokontany office as the campaign starts; Topright: a dog post-vaccination marked with a crayon; Bottomleft: a basket of puppies brought for vaccination; Bottomright: a line of owners and dogs waiting for vaccination.}}\label{figure-1.-photos-from-the-2018-campaign.-topleft-advertisement-for-the-campaign-posted-on-the-door-of-the-fokontany-office-as-the-campaign-starts-topright-a-dog-post-vaccination-marked-with-a-crayon-bottomleft-a-basket-of-puppies-brought-for-vaccination-bottomright-a-line-of-owners-and-dogs-waiting-for-vaccination.}}

\hypertarget{campaign-1}{%
\subsection{2019 Campaign}\label{campaign-1}}

For the 2019 campaign, instead of a static point campaign strategy,
vaccine vials(Rabisin) and the supplies needed to administer them
(needle, syringe, vaccination card for owners) were distributed to the
District Veterinarian, who then delivered the vaccination at no cost to
owners, but was directly compensated 1,500 Ar (\textasciitilde{} 0.40
USD) per rabies vaccine administered. The campaign lasted from September
6, 2019 to June 19, 2020, One week prior to her visiting each location,
the District Veterinarian advertised the vaccines by calling ahead to
the fokontany leaders and other officials who then advertised to their
communities, largely through word-of-mouth. For each animal vaccinated,
the District Veterinarian collected the animal's age and sex, and also
asked owners to approximate the distance they travelled to receive the
vaccination in meters. Researchers communicated with the District
Veterinarian about progress periodically throughout the campaign,
primarily through telephone calls. No other compensation or instructions
were provided, and we asked the District Veterinarian to administer as
many (or as few) as feasible or wanted. As the vaccinations were
delivered continuously, we were unable to do comparable post-vaccination
surveys.

\hypertarget{analyses}{%
\subsection{Analyses}\label{analyses}}

\hypertarget{campaign-resource-and-cost-comparisons}{%
\subsubsection{Campaign resource and cost
comparisons}\label{campaign-resource-and-cost-comparisons}}

We documented the overall costs and resources required for the two
vaccination efforts. We tracked the number of vaccination points, the
number of days over which these vaccinations occurred, and the number of
person days required overall (i.e.~the number of working people per day
over the campaign \citep{mazeri2021}), in addition to monetary costs. As
costs were incurred in both USD and Ariary, and the exchange rate
declined rapidly between 2018 and 2019, we used the midpoint between the
two years (3314 Ar to 1 USD) for the cost comparisons.

For the 2018 campaign, we broke costs down into the following
categories: direct vaccine costs (cost for vaccine, syringes, needles,
vaccination cards), supplies (livestock crayons, muzzles, gloves,
alcohol, swabs), food and lodging for NGO personnel and other
vaccinators during the campaign, personnel costs (per diems for the
district veterinarian, livestock field officers, local guides, and NGO
employees), and advertisement (posters and banners for advertising the
campaign). Foreign NGO volunteer expenses for travel to Madagascar were
not included in these costs. Vehicles and drivers are also not included
in these costs, as the drivers' time and vehicle use were donated to the
campaign by volunteers involved in the campaign. In 2019, costs were
split into two categories, direct vaccine costs (for same items as in
2018), and personnel costs (per vaccine fee paid to the district
veterinarian), and supplies (a generator and fuel for the veterinarian
to maintain the vaccine under cold chain during power outages).
Transportation costs were also not included as the District Veterinarian
used her own vehicle and vaccinated as part of their routine veterinary
service provisioning.

We used the data on owners' reported willingness to pay for vaccines to
estimate the proportional reduction in animals vaccinated as fees are
increased. We also estimated how this would impact cost per animal
vaccinated by approximating the costs for implementation (i.e.~those
costs that remain fixed) from costs incurred per animal vaccinated
(i.e.~vaccine, syringe, vaccination card, and per vaccination fee to
District Veterinarian in 2019), and calculating the balance between the
returns from owner payments (i.e.~increases in cost recovery per animal
vaccinated) vs.~decreasing numbers of animals vaccinated overall.

\hypertarget{coverage-estimates}{%
\subsubsection{Coverage estimates}\label{coverage-estimates}}

For the 2018 campaign, we used the transect data to estimate vaccination
coverage as the proportion of dogs sighted that were marked using a
binomial confidence interval at the commune level. For the 2019
campaign, we estimated vaccination coverage using human-to-dog ratios
(HDRs) and human population estimates
\citep{cleaveland2003, athingo2020}. We used a range of 8 - 25, based on
previous data from Madagascar \citep{ratsitorahina2009} and based on
recent estimates from household surveys in the Moramanga District
\citep{leblancclaireRabiesMadagascarThreepronged2019}. We set the point
estimate using an HDR of 19.5, the midpoint between the HDRs estimated
for two communities in the District by LeBlanc et al.~2019. We used
human population estimates from the 2018 national census in each commune
where the vaccinations were delivered \citep{instat2021}. Coverage was
estimated as the number of dogs vaccinated in total in that commune
divided by the estimated dog population. We used this same method for
the 2018 campaign, as well, to compare coverage estimated by the
post-vaccination transects vs.~HDRs.

\hypertarget{dog-demography}{%
\subsubsection{Dog Demography}\label{dog-demography}}

Using the age data on vaccinated animals collected during both
vaccination campaigns, we estimated the proportion of the population in
four age classes: puppies under the age of 1 year, juveniles aged 1 - 2
yrs, adults aged 2 - 6 years, and older dogs aged 6+ years based on .
With the assumption that these estimates represent the population at a
stable age distribution, we use a Leslie matrix model to estimate annual
adult survival probability and fertility using maximum likelihood
estimation \citep{Fujiwara2017}. Specifically, we assume that the number
of individuals in each age class follows a Poisson distribution with the
mean predicted by the stable age distribution from the model (the
proportion of individuals in each age class at equilibrium, equal to the
eigenvector associated with the dominant eigenvalue of the matrix
\(\nu\)) multiplied by the total number of individuals in the population
(\(N_t\)):

\[ N_{a} \sim Pois(\nu N_t) \]

We assume that all individuals older than 1 year of age reproduce, and
we do not estimate declines in fertility given the small proportion of
dogs older than age six years in the population. To get bootstrapped
estimates, we used 100 sub-sampled data sets of 1000 observations each
from the observed age data to fit the parameters, and also varied
initial values used in the optimization (N = 100 initial values sets)
for 10,000 parameter estimates total.

\hypertarget{modeling-vaccination-campaign-strategies}{%
\subsubsection{Modeling vaccination campaign
strategies}\label{modeling-vaccination-campaign-strategies}}

We used the parameter estimates from the demographic model to simulate
different vaccination strategies in a hypothetical commune with 1000
dogs. We used a discrete time age-structured model with a monthly time
step to compare three strategies:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  Annual vaccination campaigns occurring within the same month each year
  targeting dogs of all ages
\item
  Continuous vaccination of new puppies throughout the year targeting
  puppies that reach the age of 3 months
\item
  A combined approach with annual campaigns (as 1) and routine puppy
  vaccination (as 2)
\end{enumerate}

We split the dog population into puppies (\textless{} 1 year old) and
adults based on the stable age distribution estimated from the
demographic model. To estimate pup survival in year one, we took the
fertility estimates (i.e., number of new puppies per reproducing dog
observed in the pup age class) and divided by an estimate of newborn
pups per reproducing dog each year based on average litter size, average
number of litters per female per year
\citep{hampson2009, morters2014, czupryna2016}, and the proportion of
the adult population that is female (estimated from our data). We
assumed that for the annual campaign strategy, surviving vaccinated
adult dogs were revaccinated in subsequent years
\citep{delriovilas2017}, but that if a pup had been vaccinated within 9
months of the campaign, it was not revaccinated. We also assumed that
vaccine immunity lasted for a discrete period of 3 years (with
revaccination resetting immunity). A subset of parameter estimates
resulted in estimates of population decline, but based on the shape of
the age pyramid and to simulate reasonable campaign scenarios, we
filtered to parameter estimates which corresponded to positive
population growth.

\hypertarget{data-and-ethics-statement}{%
\paragraph{Data and ethics statement}\label{data-and-ethics-statement}}

All data were analysed in R version 4.0.2 (2020-06-22) \citep{R2020},
largely using the \texttt{tidyverse} package suite \citep{wickham2019}.
Geospatial data were mapped using the sf \citep{sf} package. All data
and code are available at \url{https://github.com/mrajeev08/mora_vax}.
The vaccinations were part of a public health campaign and routine
veterinary service provisioning carried out by the local veterinary
officials, the NGOs involved, and in partnership with the Ministry of
Public Health and the Department of Veterinary Services at the national
level. MDI also maintains national research permits (MICET permit:
\#130-19/MEDD/SG/DGEF/DGRNE) for its research and volunteer programs.

@@

\hypertarget{summary-of-2018-and-2019-campaigns}{%
\subsection{Summary of 2018 and 2019
campaigns}\label{summary-of-2018-and-2019-campaigns}}

During the 2018 campaign, a total of 3137 animals were vaccinated (2057
dogs and 1080 cats) in the Moramanga (urban) and Andasibe (rural)
communes over 13 days during the month of April (Table 1). We vaccinated
at 7 points in Andasibe and 14 points in Moramanga Ville. During the
2019 campaign, between September 2019 - June 2020, the District
Veterinarian vaccinated a total of 2385animals (1898 dogs and 486 cats)
over 48 days in seven communes in the Moramanga District. While more
animals were vaccinated per vaccination point and per vaccination day in
2018 compared to 2019, the number of animals vaccinated per person-day
was much higher for the 2019 campaign. More animals were vaccinated in
2018 vs.~2019, but this was largely a result of vaccinating more cats
during the 2018 campaign (Table 1).

In 2018, 15\% of dogs had a previous history of vaccination (largely in
the urban commune of Moramanga Ville) with only 7\% of dogs vaccinated
within the last year. This remained largely the same in 2019
(\textasciitilde{} 13\%), as the District Veterinarian focused their
efforts in other communes. The District Veterinarian did vaccinate 771
dogs in Moramanga Ville in 2019, and of those 24.9\% had been vaccinated
in the previous year's campaign, whereas only 4.9\% of dogs vaccinated
in all other communes had any history of previous vaccination. In
addition, in 2019, 2.1\% of animals had been spayed or neutered,
reflecting efforts by the Mad Dog Initiative to implement free spay and
neuter clinics in the district.

In 2018, 19\% of owners reported their animals were free-roaming, but
this varied by location (Table 1). In addition, while less than 19\% of
owners in Moramanga Ville reported that their animals were free roaming,
the majority of animals observed during the transects (77\%) were
observed outside of fences and not tied, and thus the majority of
animals could be classified as semi-confined in the more urban township
of Moramanga Ville, and free-roaming in the rural setting of Andasibe
(approximately 67\% of owners reported their dogs as free-roaming, Table
1). In 2019, the District Veterinarian also asked owners to approximate
how far they travelled in meters to get their animals vaccinated, and
94\% of people reportedly travelled less than 1 km to reach the
vaccination point.

\hypertarget{table-1.-summary-of-2018-and-2019-campaigns.-breakdown-of-of-animals-vaccinated-prior-vaccination-history-dog-demography-dog-ownership-daily-and-per-vaccination-rates-by-year-and-location-for-2018.-photo-credict-jochem-lastdrager-traveling-animal-doctors.}{%
\paragraph{\texorpdfstring{\textbf{Table 1. Summary of 2018 and 2019
campaigns}. \emph{Breakdown of of animals vaccinated, prior vaccination
history, dog demography, dog ownership, daily and per vaccination rates
by year and location (for 2018). Photo credict: Jochem Lastdrager,
Traveling Animal
Doctors.}}{Table 1. Summary of 2018 and 2019 campaigns. Breakdown of of animals vaccinated, prior vaccination history, dog demography, dog ownership, daily and per vaccination rates by year and location (for 2018). Photo credict: Jochem Lastdrager, Traveling Animal Doctors.}}\label{table-1.-summary-of-2018-and-2019-campaigns.-breakdown-of-of-animals-vaccinated-prior-vaccination-history-dog-demography-dog-ownership-daily-and-per-vaccination-rates-by-year-and-location-for-2018.-photo-credict-jochem-lastdrager-traveling-animal-doctors.}}

\captionsetup[table]{labelformat=empty,skip=1pt}
\begin{longtable}{lllll}
\toprule
& \multicolumn{3}{c}{2018} & \\ 
 \cmidrule(lr){2-4}
 & Andasibe & Moramanga Ville & All Communes & All Communes \\ 
\midrule
Total animals 
 vaccinated & 528 & 2609 & 3137 & 2385 \\ 
Total dogs 
 vaccinated & 254 & 1,803 & 2,057 & 1,898 \\ 
Dogs with history of vaccination & 5\% & 16\% & 15\% & 13\% \\ 
Dogs vaccinated within last year & 5\% & 7\% & 7\% & 13\% \\ 
Percent male dogs & 55\% & 56\% & 56\% & 65\% \\ 
Average dogs per owner & 0.8 & 1.1 & 1.0 & NA \\ 
Percent of owners with 
 free-roaming dogs & 67\% & 19\% & 28\% & NA \\ 
Animals vaccinated per day (total days) & 88 (6) & 372.7 (7) & 241.3 (13) & 49.7 (48) \\ 
Animals vaccinated per 
 vaccination point (total points) & 75.4 (7) & 186.4 (14) & 149.4 (21) & 37.3 (64) \\ 
Animals vaccinated per 
 person day (total person days) & 11.7 (45) & 21.6 (121) & 18.9 (166) & 49.7 (48) \\ 
\bottomrule
\end{longtable}

\hypertarget{cost-comparison-and-willingness-to-pay}{%
\subsection{Cost comparison and willingness to
pay}\label{cost-comparison-and-willingness-to-pay}}

The 2018 campaign cost more overall and per animal vaccinated than the
2019 campaign, largely due to increased personnel costs (fig 2A \& B,
Table 1). Reflecting the extra personnel necessary to run the static
point campaign, the 2018 campaign also took substantially more
person-days per animal vaccinated (Table 1). We found that charging
owners for vaccination would result in minimal cost recovery (fig 2C)
and beyond a minimal cost would actually result in increased costs per
vaccinated individual than free-of-charge campaigns. Cost recovery would
be more likely given a 2019 style strategy where the majority of the
costs are incurred on a per animal basis, compared to the 2018 campaigns
where the costs are largely due to setting up the static point
vaccination stations (fig 4A). More importantly, in all cases, even a
nominal fee would significantly reduce numbers of dogs vaccinated and
thus vaccination coverage, particularly in the rural commune of Andasibe
(fig 2D).

\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig2}

\hypertarget{figure-2.-comparing-campaign-costs-and-willingness-to-pay.-a-vaccine-costs-broken-down-by-category-and-by-year-colors.-b-overall-cost-per-animal-vaccinated-for-the-two-campaign-years-split-by-direct-costs-of-vaccination-per-animal-i.e.-vaccine-vaccination-card-syringes-and-baseline-implementation-costs-i.e.-personnel-supplies-subsistence-costs-for-vaccinators-during-the-campaign.-c-estimated-cost-per-animal-vaccinated-under-a-willingness-to-pay-model-for-two-campaigns-examining-increasing-costs-charged-to-the-owner-with-estimated-costs-declining-due-to-cost-recovery-through-owner-payments-and-then-peaking-once-owners-report-to-be-no-longer-willing-to-pay-for-the-vaccine.-d-the-percent-reduction-in-number-of-animals-vaccinated-given-owners-willingness-to-pay.-the-curves-in-c-and-d-are-shown-based-on-the-overall-responses-on-willingness-to-pay-solid-line-from-both-moramanga-and-andasibe-and-the-responses-split-by-commune-dashed-and-dotted-lines.}{%
\paragraph{\texorpdfstring{\textbf{Figure 2. Comparing campaign costs
and willingness to pay}. \emph{A) Vaccine costs broken down by category
and by year (colors). B) Overall cost per animal vaccinated for the two
campaign years, split by direct costs of vaccination per animal
(i.e.~vaccine, vaccination card, syringes) and baseline implementation
costs (i.e.~personnel, supplies, subsistence costs for vaccinators
during the campaign). C) Estimated cost per animal vaccinated under a
willingness to pay model for two campaigns examining increasing costs
charged to the owner, with estimated costs declining due to cost
recovery through owner payments and then peaking once owners report to
be no longer willing to pay for the vaccine. D) The percent reduction in
number of animals vaccinated given owners willingness to pay. The curves
in C and D are shown based on the overall responses on willingness to
pay (solid line) from both Moramanga and Andasibe, and the responses
split by Commune (dashed and dotted
lines).}}{Figure 2. Comparing campaign costs and willingness to pay. A) Vaccine costs broken down by category and by year (colors). B) Overall cost per animal vaccinated for the two campaign years, split by direct costs of vaccination per animal (i.e.~vaccine, vaccination card, syringes) and baseline implementation costs (i.e.~personnel, supplies, subsistence costs for vaccinators during the campaign). C) Estimated cost per animal vaccinated under a willingness to pay model for two campaigns examining increasing costs charged to the owner, with estimated costs declining due to cost recovery through owner payments and then peaking once owners report to be no longer willing to pay for the vaccine. D) The percent reduction in number of animals vaccinated given owners willingness to pay. The curves in C and D are shown based on the overall responses on willingness to pay (solid line) from both Moramanga and Andasibe, and the responses split by Commune (dashed and dotted lines).}}\label{figure-2.-comparing-campaign-costs-and-willingness-to-pay.-a-vaccine-costs-broken-down-by-category-and-by-year-colors.-b-overall-cost-per-animal-vaccinated-for-the-two-campaign-years-split-by-direct-costs-of-vaccination-per-animal-i.e.-vaccine-vaccination-card-syringes-and-baseline-implementation-costs-i.e.-personnel-supplies-subsistence-costs-for-vaccinators-during-the-campaign.-c-estimated-cost-per-animal-vaccinated-under-a-willingness-to-pay-model-for-two-campaigns-examining-increasing-costs-charged-to-the-owner-with-estimated-costs-declining-due-to-cost-recovery-through-owner-payments-and-then-peaking-once-owners-report-to-be-no-longer-willing-to-pay-for-the-vaccine.-d-the-percent-reduction-in-number-of-animals-vaccinated-given-owners-willingness-to-pay.-the-curves-in-c-and-d-are-shown-based-on-the-overall-responses-on-willingness-to-pay-solid-line-from-both-moramanga-and-andasibe-and-the-responses-split-by-commune-dashed-and-dotted-lines.}}

\hypertarget{comparing-campaign-coverage-estimates}{%
\subsection{Comparing campaign coverage
estimates}\label{comparing-campaign-coverage-estimates}}
